A Parallel Geometric Multigrid Algorithm

### 2014-09-24 branches/testcharge/htdocs/multigrid.tex
sync merge trunk to testcharge
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2478 40876441-3496-0410-a613-a2fca35b2ae2

\begin{eqnarray*}
k&\dots&\mbox{grid level with $k=1$ being the coarsest grid}\\
\Omega_k &\dots& \mbox{regular grid with spacing $h(k)$}\\
\partial \Omega_k &\dots& \mbox{boundary of computational domain}\\
m&\dots& \mbox{iteration index}\\
u_k&\dots& \mbox{exact solution on $\Omega_k$}\\ 
v_k^m&\dots& \mbox{approximate solution on $\Omega_k$ after $m$ iterations}\\
e_k^m&\dots& \mbox{error on $\Omega_k$ after $m$ iterations}\\ 
R_h^H, R_k^{k-1}&\dots& \mbox{restriction operator: mapping functions on grid level $k$ with} \\
&& \mbox{spacing $h$ to functions on the next coarser level $k-1$ with spacing $H=2h$}\\ 
I_H^h, I_{k-1}^k&\dots& \mbox{interpolation operator: mapping functions on grid level $k-1$ with }\\
&&\mbox{spacing $H=2h$ to functions on the next finer level $k$ with spacing $h$}\\ 
\end{eqnarray*}

@@

Probably the easiest way to deal with this issue is to truncate the multigrid hierarchy so that there are always plenty of grid points on each process. This can have a detrimental effect on the overall-convergence of the algorithm as the lowest grid level is probably still too fine to achieve rapid convergence.

Another strategy is to use a fast, parallel iterative solver on the coarstest grid level like conjugate-gradient solver for instance.

It is also possible to gather grid points that are initially allocated on different processes, solve only on a subset of processes with the others idling and scatter afterwards. Alternatively the data from all processes can be distributed over the entire topology by means of an MPI All-to-All call. The identical calculation can then be performed simulataneously on all processes, making further communication after each iteration step unnecessary. The grid can be also further coarsed down to the level where a direct solutin is possible.

@@

\begin{algorithm}
\caption{Multigrid V-cycle for one iteration step $v_k^m\rightarrow v_k^{m+1}$}
\begin{algorithmic} 
\STATE{}
\STATE{1. Presmoothing:}
\FOR{$\nu_1$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k v_k = f_k$ on $\partial\Omega_k$ with initial guess $v_k$}
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k v_k = f_k$ on $\Omega_k\setminus \partial\Omega_k$  with initial guess $v_k$}
\ENDFOR
\STATE{}
\STATE{2. Compute residual:}
\STATE{$\triangleright$ Compute $r_k=f_k-A_k v_k \rightarrow f_k$}
\STATE{}
\STATE{3. Restrict residual:}
\STATE{$\triangleright$ Compute $f_{k-1}=R^{k-1}_k f_k$}
\STATE{}
\STATE{4. Coarse solve:}
\IF{$k-1>1$}
\FOR{$\nu_1$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_{k-1}v_{k-1}=f_{k-1}$ on $\partial\Omega_{k-1}$ with initial guess $v_{k-1}=0$ }
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_{k-1}v_{k-1}=f_{k-1}$ on $\Omega_{k-1}\setminus \partial\Omega_{k-1}$ with initial guess $v_{k-1}=0$ }
\ENDFOR
\STATE{$\triangleright$ Set $k-1 \rightarrow k$; goto step 2.}
\ENDIF
\STATE{}
\IF{$k-1=1$}
\STATE{$\triangleright$ Scatter all data via MPI All-to-All call}
\STATE{$\triangleright$ On each process solve $A_1 v_1=f_1$ on entire domain $\Omega_1$, either exactly or using a fast iterative solver with inital guess $v_1=0$} 
\ENDIF
\STATE{}
\STATE{5. Apply coarse grid correction to inital guess:}
\STATE{$\triangleright$ Compute $\hat{v}_k=v_k + I_{k-1}^k v_{k-1}$ on $\Omega_k$}
\STATE{}
\STATE{6. Postsmoothing:}
\FOR{$\nu_2$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k \hat{v}_k= f_k$ on $\partial\Omega_k \rightarrow v_k$ new approximate solution}
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k \hat{v}_k= f_k$ on $\Omega_k\setminus\partial\Omega_k \rightarrow v_k$ new approximate solution}
\ENDFOR
\STATE{}
\end{algorithmic}
\end{algorithm}

@@

The following remarks are relevant for the branch of the code living
in the \texttt{testcharge} directory, where we have SOR (both uniform
permitivity and heterogenous permitivity) and should shortly have a
Fourier solver (uniform permeativity) available.

@@

The existing code has a single coordinaite system which corresponds
to the fine grid $\Omega_h$ in the multipgrid picture. Details of the
coordinate system (global and local extent; MPI communicators; 3d-to-1d
indexing functions, etc) are encapsulated in \texttt{coords.c}.
\begin{enumerate}
\item
As a first step towards a multigrid implementation, it would be
desirable to ``objectify'' formally a \texttt{coods\_t} data type.
This would be explicitly referenced in all existing fine grid
contexts (and is desirable on other grounds as well).
\item
For multigrid, one could then instantiate one \texttt{coords\_t}
object for each level of the grid $\Omega_k$ to allow corresponding
grid operations at a given level.
\item
A second option would be to retain a single \texttt{coords\_t} object,
but add functionality to recognise a ggrid level $k$ and perform the
necessary arithmetic/logic assuning uniform refinement. (This is
probably a less elegant and flexible solution, if more compact.)
\end{enumerate}

@@

Each level of the grid $\Omega_k$ requires coordinate information,
but also local storage for three-dimensional field quantities, eg.,
$f_k(x,y,z)$. Local field quantities are expected to require a halo
region of extent \texttt{nhalo} cf. existing field quantities on the
fine grid. This allows interpolation and restriction operators to be
evalautated locally.

>>>>

\begin{eqnarray*}
k&\dots&\mbox{grid level $1\le k \le k_{max}$ with $k=1$ being the coarsest grid}\\
\Omega_k &\dots& \mbox{regular grid with spacing $h(k)$}\\
\partial \Omega_k &\dots& \mbox{boundary of computational domain}\\
m&\dots& \mbox{iteration index}\\
u_k&\dots& \mbox{exact solution on $\Omega_k$}\\ 
v_k^m&\dots& \mbox{approximate solution on $\Omega_k$ after $m$ iterations}\\
e_k^m&\dots& \mbox{error on $\Omega_k$ after $m$ iterations}\\ 
R_h^H, R_k^{k-1}&\dots& \mbox{restriction operator: mapping functions on grid level $k$ with} \\
&& \mbox{spacing $h$ to functions on the next coarser level $k-1$ with spacing $H=2h$}\\ 
I_H^h, I_{k-1}^k&\dots& \mbox{interpolation operator: mapping functions on grid level $k-1$ with }\\
&&\mbox{spacing $H=2h$ to functions on the next finer level $k$ with spacing $h$}\\ 
\end{eqnarray*}

@@

Probably the easiest way to deal with this issue is to truncate the multigrid hierarchy so that there are always plenty of grid points on each process. This can have a detrimental effect on the overall-convergence of the algorithm as the lowest grid level is probably still too fine to achieve rapid convergence. It is also possible to use a fast, parallel iterative solver like conjugate-gradient on the coarstest grid.

In a first approach we will opt for another strategy. On a coarse grid the data from all individual processes can be distributed over the entire topology by using an All-to-All MPI-call. An exactly identical calculation is then performed on all processes making communication after each iteration step unnecessary. The grid can be also further coarsened down to the level where a direct solution becomes feasible.

@@

\begin{algorithm}[H]
\caption{Multigrid V-cycle for one iteration step $v_k^m\rightarrow v_k^{m+1}$}
\begin{algorithmic} 
\STATE{}
\STATE{1. Presmoothing:}
\FOR{$\nu_1$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k v_k = f_k$ on $\partial\Omega_k$ with initial guess $v_k$}
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k v_k = f_k$ on $\Omega_k\setminus \partial\Omega_k$  with initial guess $v_k$}
\ENDFOR
\STATE{}
\STATE{2. Compute residual:}
\STATE{$\triangleright$ Compute $r_k=f_k-A_k v_k \rightarrow f_k$}
\STATE{}
\STATE{3. Restrict residual:}
\STATE{$\triangleright$ Compute $f_{k-1}=R^{k-1}_k f_k$}
\STATE{}
\STATE{4. Coarse solve:}
\IF{$k-1>1$}
\FOR{$\nu_1$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_{k-1}v_{k-1}=f_{k-1}$ on $\partial\Omega_{k-1}$ with initial guess $v_{k-1}=0$ }
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_{k-1}v_{k-1}=f_{k-1}$ on $\Omega_{k-1}\setminus \partial\Omega_{k-1}$ with initial guess $v_{k-1}=0$ }
\ENDFOR
\STATE{$\triangleright$ Set $k-1 \rightarrow k$; goto step 2.}
\ENDIF
\STATE{}
\IF{$k-1=1$}
\STATE{$\triangleright$ Scatter all data via MPI All-to-All call}
\STATE{$\triangleright$ On each process solve $A_1 v_1=f_1$ on entire domain $\Omega_1$, either exactly or using a fast iterative solver with inital guess $v_1=0$} 
\ENDIF
\STATE{}
\STATE{5. Apply coarse grid correction to inital guess:}
\STATE{$\triangleright$ Compute $\hat{v}_k=v_k + I_{k-1}^k v_{k-1}$ on $\Omega_k$}
\STATE{}
\STATE{6. Postsmoothing:}
\FOR{$\nu_2$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k \hat{v}_k= f_k$ on $\partial\Omega_k \rightarrow v_k$ new approximate solution}
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k \hat{v}_k= f_k$ on $\Omega_k\setminus\partial\Omega_k \rightarrow v_k$ new approximate solution}
\ENDFOR
\STATE{}
\end{algorithmic}
\end{algorithm}

@@

The following remarks are relevant for the branch of the code living
in the \texttt{testcharge} directory, where we have SOR (both uniform
permittivity and heterogenous permittivity) and should shortly have a
Fourier solver (uniform permittivity) available.

@@

The existing code has a single coordinate system which corresponds
to the fine grid $\Omega_h$ in the multigrid picture. Details of the
coordinate system (global and local extent; MPI communicators; 3d-to-1d
indexing functions, etc) are encapsulated in \texttt{coords.c}.
\begin{enumerate}
\item
As a first step towards a multigrid implementation, it would be
desirable to ``objectify'' formally a \texttt{coods\_t} data type.
This would be explicitly referenced in all existing fine grid
contexts (and is desirable on other grounds as well).
\item
For multigrid, one could then instantiate one \texttt{coords\_t}
object for each level of the grid $\Omega_k$ to allow corresponding
grid operations at a given level.
\item
A second option would be to retain a single \texttt{coords\_t} object,
but add functionality to recognise a grid level $k$ and perform the
necessary arithmetic/logic assuming uniform refinement. (This is
probably a less elegant and flexible solution, if more compact.)
\end{enumerate}

@@

Each level of the grid $\Omega_k$ requires coordinate information,
but also local storage for three-dimensional field quantities, eg.,
$f_k(x,y,z)$. Local field quantities are expected to require a halo
region of extent \texttt{nhalo} cf. existing field quantities on the
fine grid. This allows interpolation and restriction operators to be
evaluated locally.
